{\small} High Performance P$^{3}$M N-body code: CUBEP3M

### 2012-06-12 Extendedpp.tex
Included force tests over 10 steps
added a rebined force plot to show how the mean behaves

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
 plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
 which are now at least 5 time smaller than in the default P$^{3}$M algorithm.

 \begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
  \caption{Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_ppext0}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.. 
    \label{fig:den_force_ppext6}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
  \caption{Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
  in a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
  This was obtained over a single time step.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. {\bf (y-axis?)}}
    \label{fig:MassFunction_extpp}
  \end{center}
\end{figure}

>>>>

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
 plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to six layers of fine cells. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
 which are now at least 5 time smaller than in the default P$^{3}$M algorithm.

 \begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
%    \label{fig:den_force_ppext6}}
%\end{center}
%\end{figure}
%\begin{figure}%[ht]
%  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
    \caption{({\it left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
  %\caption{
  ({\it right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
  calculated from a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
  This was also obtained over a single time step.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure*}

@@

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. 
  The smallest haloes have a mass equivalent to 20 particles.{\bf (y-axis?)}}
    \label{fig:MassFunction_extpp}
  \end{center}
\end{figure}

### 2012-06-12 Halos.tex
added a rebined force plot to show how the mean behaves

A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
contains 200-2000 particles and so on. Comparisons with the linear predictions of \cite{Sheth_et_al} shows
results consistent findings of \cite{Thinker}, namely that low-mass bias is over-predicted and high-mass bias is under-predicted.

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/bias.eps}
  \end{center}
  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2,000 and so on.
  The dashed lines are linear predictions from \protect \cite{Sheth_et_al}.
    \label{fig:halo}}
\end{figure}

>>>>

A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
contains 200-2000 particles and so on. Comparisons with the linear predictions of \citet{2001MNRAS.323....1S} shows
results consistent the known flaws of the model, namely that low-mass bias is over-predicted and high-mass bias is under-predicted \citep{2010ApJ...724..878T}.

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/bias.eps}
  \end{center}
  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2000 and so on.
  The dashed lines are linear predictions from \protect \citet{2001MNRAS.323....1S}.
    \label{fig:halo}}
\end{figure}

### 2012-06-12 Systematics.tex
added a rebined force plot to show how the mean behaves

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed in the initial conditions. {\b (I should add a vertical line at the k of the coarse mesh for visualization purposes.)}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

To illustrate how well the random offset performs, we present in Fig. \ref{fig:den_force_N10}
the impact of averaging over 10 time steps. We recast the same force test in which we dig a hole, 
with the addition that we apply a random offset the whole particle population at the end of the calculation.
We repeat this operation ten times, always extracting the same particle, and average the mean force and the deviation.
When compared with a single time step, we see that the scatter has reduced by almost a factor of three in the radial direction, 
while the transverse deviation shows a milder improvement.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
  \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
    ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
    The left panels are calculated from a single time step, while the right panels show the average over time time steps.
    \label{fig:den_force_N10}}
\end{center}
\end{figure*}

>>>>

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
The fluctuations at low-$k$ come from the white noise imposed in the initial conditions. The vertical line corresponds to the scale of the coarse mesh.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}. 
  The vertical line corresponds to the scale of the coarse mesh.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

To illustrate how well the random offset performs, we present in Fig. \ref{fig:den_force_N10}
the impact of averaging over 10 time steps. We recast the same force test in which we dig a hole, 
with the addition that we apply a random offset the whole particle population at the end of the calculation.
We repeat this operation ten times, always extracting the same particle, and average the mean force and the deviation.
When compared with a single time step, we see that the scatter has reduced by almost a factor of three in the radial direction, 
while the transverse deviation shows a milder improvement. When rebining these results, we observe that the mean mesh force
calculation agrees well down to the grid cell, as seen in Fig. \ref{fig:F_rebin}. For this calculation, the hole was not part of a very dense halo,
hence most of the force it experiences comes from the mesh.

\begin{figure*}%[ht]
  \begin{center}
     \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
   \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
  \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
    ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
    The left panels are calculated from a single time step, while the right panels show the average over time time steps.
    \label{fig:den_force_N10}}
\end{center}
\end{figure*}

\begin{figure}%[ht]
  \begin{center}
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=10_rebin.eps}
  \caption{ Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$, averaged over 10 time steps.
    Results are rebined to show how well the mean value follows the thoretical curve down to the grid size.
    \label{fig:F_rebin}}
\end{center}
\end{figure}
